The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2016-11-28 manuscript/manuscript.Rnw
fix syncytin typo
cite hutchinson for maternal immune response

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
can be attributed to the lack of adequate invasion of maternal tissue,
and is noted by the onset of hypertension and proteinuria (or in the
absence of proteinuria, thrombocytopenia, impaired liver function,
renal insufficiency, pulmonary edema, or cerebral or visual
disturbances) %
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
pre-eclampsia can develop into eclampsia, a more severe condition
resulting in seizures during pregnancy and death. Lack of invasion in
pre-eclampsia may be attributed to the inefficiency of growth
regulators as well as the inability to sufficiently suppress the
maternal immune response. %
Although there are speculative cases of eclampsia in great apes, 
%\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
it is exceedingly rare in these species
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.

@@

@@

While Syncitin-1 and -2 are implicated in the formation of the
syncytiotrophoblast in
\human~\cite{Rawn.ea2008:evolutionregulationandfunction} and homologs
of \textit{ERVFRD-1} and \textit{ERVW-1} are widely expressed in most
of the placentas analyzed here, they are not expressed in \bonobo,
\sheep, and \dog, even though these mammals form
syncytiotrophoblasts~\cite{Mess.ea2006:Evolutionarytransformationsoffetal}.
Because the formation of syncytiotrophoblast is the stem state of
Eutheria, Afrotheria, Euarchontoglires, and
Laurasiatheria~\cite{Mess.ea2006:Evolutionarytransformationsoffetal},
the gene(s) responsible for cytotrophoblast fusion to form
syncytiotrophoblast is likely to be highly expressed in the core
placenta. \textit{ANXA5} is known to be expressed at multiple stages
of placentation, and is capable of repairing membrane defects and
promoting the fusion of
cells~\cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing} in
the presence of Ca$^{2+}$. While there are likely other genes which
are responsible for the positioning of cytotrophoblast cells in close
proximity to start the fusion process, \textit{ANXA5} is likely to be
essential for the completion of fusion and the maintenance of membrane
repair during the massive flux of nutrients and waste in the placenta.

>>>>

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
can be attributed to the lack of adequate invasion of maternal tissue,
and is noted by the onset of hypertension and proteinuria (or in the
absence of proteinuria, thrombocytopenia, impaired liver function,
renal insufficiency, pulmonary edema, or cerebral or visual
disturbances) %
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
pre-eclampsia can develop into eclampsia, a more severe condition
resulting in seizures during pregnancy and death. Lack of invasion in
pre-eclampsia may be attributed to the inefficiency of growth
regulators as well as the inability to sufficiently suppress the
maternal immune response~\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. %
Although there are speculative cases of eclampsia in great apes, 
%\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
it is exceedingly rare in these species
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.

@@

\DEW{Why?}

@@

While Syncytin-1 and -2 are implicated in the formation of the
syncytiotrophoblast in
\human~\cite{Rawn.ea2008:evolutionregulationandfunction} and homologs
of \textit{ERVFRD-1} and \textit{ERVW-1} are widely expressed in most
of the placentas analyzed here, they are not expressed in \bonobo,
\sheep, and \dog, even though these mammals form
syncytiotrophoblasts~\cite{Mess.ea2006:Evolutionarytransformationsoffetal}.
Because the formation of syncytiotrophoblast is the stem state of
Eutheria, Afrotheria, Euarchontoglires, and
Laurasiatheria~\cite{Mess.ea2006:Evolutionarytransformationsoffetal},
the gene(s) responsible for cytotrophoblast fusion to form
syncytiotrophoblast is likely to be highly expressed in the core
placenta. \textit{ANXA5} is known to be expressed at multiple stages
of placentation, and is capable of repairing membrane defects and
promoting the fusion of
cells~\cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing} in
the presence of Ca$^{2+}$. While there are likely other genes which
are responsible for the positioning of cytotrophoblast cells in close
proximity to start the fusion process, \textit{ANXA5} is likely to be
essential for the completion of fusion and the maintenance of membrane
repair during the massive flux of nutrients and waste in the placenta.
